---
title: "Lesson11"
author: "Vicki Hertzberg"
date: "3/29/2017"
output: html_document
---

# Analyzing Microbiome Data

Most of you are here today because you have an interest in the microbiome, that "new organ" comprising the trillions of microbes that live in and on us. Today we are going to talk about how to identify these organisms isolated from a sample by sequencing a portion of the genetic material present, lining these up into counts of organisms, then using these data to summarize the microbes present in numeric or graphical fashion.

Before we get to that I just have one short message: (https://www.youtube.com/watch?v=htbeJhtFAXw)[https://www.youtube.com/watch?v=htbeJhtFAXw]

We will proceed as follows:

2. We will by briefly discussing the concept of metagenomics in general, and the 16S region of the bacterial genome in particular.

1. We will next discuss sequencing the .fastq file format. We are talking about .fastq files to give you an idea of what is coming out of the sequencer, why there is uncertainty, and why this is all so darned complicated.

3. We will walk through an example of processing a set of forward and reverse .fastq files to come up with something akin to an OTU table.

4. We will next use our "OTU table" together with data about our samples to produce numerical and visual descriptive summaries of our samples. 

5. We will end by again using this "OTU table" and our sample data to develop functional profiles of our samples.

## Metagenomics

_Genomics, Proteomics, Transcriptomics_

- These focus on the *whole* of a *single* individual organism.
- But organisms live in communities and interact.
- To study a community, we need to study across the genomes of the individual organisms comprising it.

_Metagenomics_

- The study of metagenomes, genetic material sampled directly from a community in its natural environment.
- Obviates the need for isolation and lab cultivation.

There are several methods for characterizing metagenomes. Some of these involve sequencing amplified markers, "amplicons" of eukaryotic genomes. For bacteria and archaeae, these are typically highly conserved small subunits of the gene that encodes for ribosomal RNA, in particular the 5S, 16S, 18Sm or 23S regions. For instance, the 16S rRNA unit, about 1500 base pairs in length, has 9 hypervariable regions. For fungi, the ITS unit serves this purpose. You can design an experiment so that you target one of those regions, or you can sequence the whole 16S region. Your choice of primers (and budget) will determine how much you sequence. 

Alternatively you can perform whole genome sequencing (WGS), sometimes also called "shotgun" sequencing, to determine the whole bacterial genome. WGS is about 10 times more  expensive then 16S sequencing.

_16S rRNA_

Universal PCR primers are used to amplify this region. Primers have been designed to match highly conserved regions of this subunit. A 2-D representation of the 16S region for the *Thermus thermophilus*, an extremophile bacteria, can be found at (http://rna.ucsc.edu/rnacenter/images/figs/thermus_16s_2ndry.jpg)[http://rna.ucsc.edu/rnacenter/images/figs/thermus_16s_2ndry.jpg]

_V4 Hypervariable Region_

Universal primers:

- Forward (515F): GTGCCAGCMGCCGCGGTAA
- Reverse (805R): GGACTACHVGGGTWTCTAA
- These will produce 288 - 290 bp amplicons covering the V4 region

Sequences determined by Illumina MiSeq protocol

- Should get nearly complete overlap of forward and reverse reads.
- These can be merged to form a high quality consense base call at each position.
- Then classify the merged reads.

_Wait! I thought that the only letters allowed in the sequences are A, C, G, or T._ Yet I see M, H, V, and W in those primer sequences. What's up with that?

It turns out that the Illumina is smart enough to know what it doesn't know. So if it knows the base, it calls it that way. OTOH, if it can only resolve the base call down to, say, the choice of 2 bases, it uses these other letters to designate that situation. These are called the IUPAC Ambiguity Codes, and you can see them here: (http://droog.gs.washington.edu/parc/images/iupac.html)[http://droog.gs.washington.edu/parc/images/iupac.html]. There is another IUPAC code not included on the table, and that is "-", which is used to indicate a gap, an often useful indicator for alignments.


## Sequencing File Format

In the files produced by the sequencer, every read represents an independent copy of source DNA in your sample. When the target material is sequenced, there are two main considerations: sequencing *breadth* and sequencing *depth*.

- _Breadth_ refers to the extent to which you sequence the entire genome present. You want to be sure that you have sequence information for all areas of the target.
- _Depth_ refers to how many reads on average cover each base pair of your target. This is also sometimes referred to as "coverage."

- In 16s rRNA amplicon sequencing, the primers that you use determine breadth. 
- If you don't have sufficient depth you may end up with incomplete or inconclusive results. OTOH, oversequencing raises costs. 



### Illumina FASTQ Files

#### File Naming Conventions:

- NA10831_ATCACG_L002_R1_001.fastq.gz 
- FA1_S1_L001_R1_001.fastq.gz 
- Sample_Barcode/Index_Lane_Read#_Set#.fastq.gz

#### Sequence Identifiers

- \@EAS139:136:FC706VJ:2:5:1000:12850 1:N:18:ATCACG 
- \@M00763:36:000000000-A8T0A:1:1101:14740:1627 1:N:0:1
- \@Instrument : Run# : FlowcellID : Lane : Tile : X : Y Read : Filtered : Control# : Barcode/Index

#### FASTQ File Format

| 4 lines per read as follows: |
|------------------------------|
| \@sequence_id                |
| sequence                     |
| \+                           |
| quality                      |
|------------------------------|



#### Example
 

- Line1: \@M00763:36:000000000-A8T0A:1:1101:14740:1627 1:N:0:1

- Line2: CCTACGGGAGGCAGCAGTGGGGAATATTGCACAATGGGGGAAACCCTGATGC AGCGACGCCGCGTGAGTGAAGAAGTATCTCGGTATGTAAAGCTCTATCAGCA GGAAAGATAATGACGGTACCTGACTAAGAAGCCCCGGCTAACTACGTGCCAG CAGCCGCGGTAATACGTAGGGGGCAAGCGTTATCCGGATTTACTGGGTGTAA AGGGAGCGTAGACGGCAGCGCAAGTCTGGAGTGAAATGCCGGGGCCCAACCC 
CGGCCCTGCTTTGGAACCCGTCCCGCTCCAGTGCGGGCGGG

- Line3: \+ 

- Line 4: 88CCCGDBAF)===CEFFGGGG>GGGGGGCCFGGGGGDFGGGGDCFGGGFED CFG:\@CFCGGGGGGG?FFG9FFFGG9ECEFGGGDFGGGFFEFAFAFFEFECE F\@4AFD85CFFAA?7+C\@FFF<,A?,,,,,,AFFF77BFC,8>,>8D\@FFFF G,ACGGGCFG>\*57;\*6=C58:?<)9?:=:C\*;;\@C?3977\@C7E\*;29>/= +2\*\*)75):17)8\@EE3>D59>)>).)61)4>(6\*+/)\@F63639993??D1 :0)((,((.(.+)(()(-(\*-(-((-,,(.(.)),(-0)))

#### Quality Scores

The 4th line consists of the Phred scores for each base call. Each character is associated with a value between 0 and 40, and these are called the Quality scores or Q scores. The coding scheme can be found at (https://support.illumina.com/help/BaseSpace_OLH_009008/Content/Source/Informatics/BS/QualityScoreEncoding_swBS.htm)[https://support.illumina.com/help/BaseSpace_OLH_009008/Content/Source/Informatics/BS/QualityScoreEncoding_swBS.htm]  

The relationship is as follows:

\begin{equation}
Q=-10*log(p)
\end{equation}

Or you can solve for p as follows:

\begin{equation}
p=10^{-Q/10}
\end{equation}

You can see from the table that letters have high Quality Scores. Let's look at the Phred score of I, which has a Q value of 40. 

|                |
|----------------|
| Q=40           |
| -Q/10 = -4     |
| p = $10^{-4}$|
| p = 0.0001     |


In contrast, consider when the Phred score is "+", the Q value is 10, and p = 0.10.

In summary, if you are looking at the .fastq file (which you can open with a text editor), letters are good, numbers are medium, and most anything else is of lesser quality.

You want to know this because you may want to trim your sequences. Sometimes when a sequence is read, the first 10 or so sequences are of lesser quality. Also, reads deteriorate in quality towards the end of a sequence. Finally, forward reads are generally of better quality than their corresponding reverse reads.

### Processing Sequencing Data: Overview

1. Filter - remove low-quality reads and non-target sequences
2. Trim - prune low-quality ends
3. Assembly - correct overlapping bases
4. Aggregate - combine similar reads
5. Chimeras - remove chimeric reads

## Processing Sequencing Data with the DADA2 Pipeline

Today we are going to go through the process of "feeding" a set of Illumina-sequenced paired-end .fastq files into the `dada2` package, with the goal of deriving a _sequence table_ in the end. The term "dada" stands for Divisive Amplicon Denoising Algorithm. This R package example uses 4 dada steps, 2 for the forward reads and 2 for the reverse reads, hence "dada2."

A sequence table is akin to the ubiquitous OTU table, only it is at a much higher resolution.  OTUs, or operational taxonomic units, is an operational term defined as sets of individual organisms that have been put together in groups according to their DNA similarity. OTUs are often used as proxies for "species". "Similarity" in this context is defined as differing by no more than a fixed dissimilarity threshold, typically set at 3%. But we are going to do better than 3% dissimilarity; we are headed to 100%. 

Why isn't 97% good enough?

1. It still is low resolution, fiving genus level identification at best
2. It tends to have a high false positive rate, with the number of OTUs much greater than actual richness.
3. As you get more samples, the processing time scales up super-linearly.

We are using `dada2` because it infers sample sequences exactly, resolving differences of as little as one (1) nucleotide. Not only does `dada2` provide higher resolution, it also performs better than two other major players in this field: mothur and QIIME (v1). But note that QIIME v2.1 now has a DADA2 step built into it, so it should now compete well with this package. 

For the moment, however, we are going to continue to use `dada2`, well, because. Because we are now expert R, RStudio, git and github users, that's why!!!

#### First Note of Importance

Our "input" into the package is a set of Illumina-sequenced paired-end .fastq files that have been demultiplexed (i.e., split by sample) and _from which the barcodes/adapters/primers have already been removed._  In this toy dataset all of these barcoode/adapters/primers have been removed. But when it comes to processing your own data, you will have to check this before you proceed. 

My personal experience to with two different labs is that you will receive demultiplexed files with the barcodes removed, *but* the primers have not been removed. You will need to know that the primers are so that you can figure out the lengths. You can trim before hand with a program such as TRIMMOMATIC (http://www.usadellab.org/cms/?page=trimmomatic)[http://www.usadellab.org/cms/?page=trimmomatic]. This way, when we display the quality plots later you will be seeing the quality of the "meat" of your data, instead of having some distraction with the quality of the sequences including the primers. 

You can also trim later by setting an option. I will show you (outside of an R chunk of course, because these data have already had the primers trimmed) how to do that.

#### Second Note of Importance

This pipeline also assumes that if you have paired reads, that the forward and reverse .fastq files contain the reads in matching order.

If this is not true, you will need to remedy this before proceeding. Please see the `dada2` FAQ for suggestions: (http://benjjneb.github.io/dada2/faq.html#what-if-my-forward-and-reverse-reads-arent-in-matching-order)[http://benjjneb.github.io/dada2/faq.html#what-if-my-forward-and-reverse-reads-arent-in-matching-order].

##### End of Notes of Importance. For the Moment.


Since I am moving my processing from my MacBook Air to a brand new souped-up MacBook Pro, I needed to install the dada2, ShortRead, and ggplot2 packages from Bioconductor. All of you should have already have these installed. Since I have now installed everything, I am now going to load those libraries and make sure that the package versions are correct. I am comparing against the `dada2` pipeline tutorial, which can be found at (http://benjjneb.github.io/dada2/tutorial.html)[http://benjjneb.github.io/dada2/tutorial.html]. In fact, most of this comes from that tutorial, which was written by Ben Callahan, who is the author of `dada2`. Note that as of today (3/24/2017), the version numbers that I will display from the R chunk below will differ slightly from those in the online tutorial. But mine are newer versions, and that should be ok. If you have older versions, you might be in trouble.

So let's start now.

### Are We Ready?

Let's check that all is ready.

```{r}
library(dada2); packageVersion("dada2")
library(ShortRead); packageVersion("ShortRead")
library(ggplot2); packageVersion("ggplot2")
```

I have also downloaded the file used in the Mothur MiSeq SOP, as well as two RDP reference files. The Mothur MiSeq files contain data from an experiment in which the V4 region of the 16S rRNA gene in mice feces was sequenced.  You will have to change the path in the next chunk to the path to where your files sit. Also if you are on a Windows machine, this will also look different. Let's make sure they are all in the proper place on my machine:

```{r}
# Set the path to the data files
path <- "~/Documents/NRSG_741/MiSeqData/MiSeq_SOP"
fileNames <- list.files(path)
fileNames
```

OK, I see 38 .fastq files and the two RDP files. With the exception of a file named "filtered" (which the `dada2` tutorial lists but we do not) and the two RDP files (which we list but the `dada2` tutorial does not), we agree. The file named "filtered" will be created in another couple of steps, so we are not going to worry about that.

### Filter and Trim

So now we are ready to use the `dada2` pipeline. We will first read in the names of the .fastq files. Then we will manipulate those names as character variables, using regular expressions to create lists of the forward and reverse read .fastq files in *matched* order.

```{r}

# Read in the names of the .fastq files

fastqFiles <- fileNames[grepl(".fastq$", fileNames)]

# Sort so that forward / reverse reads are in the same order

fastqFiles <- sort(fastqFiles)

# Get just the forward read files

fileNameForwards <- fastqFiles[grepl("_R1", fastqFiles)]

# Get just the reverse read files

fileNameReverses <- fastqFiles[grepl("_R2", fastqFiles)]

# Get the sample names, assuming file naming convention is: SAMPLENAME_XXX.fastq

sample.names <- sapply(strsplit(fileNameForwards, "_"), `[`, 1)

# Specify the full file path to the forward and reverse read.s

fileNameForwards <- file.path(path, fileNameForwards)
fileNameReverses <- file.path(path, fileNameReverses)
  
  




```

#### Important Note 3

If you are using this workflow with your own data, you will probably need to modify the R chunk above, especially the assignment of sample names to the variable `sample.names`.

#### End of Note

### Quality Profiles of the Reads

One of the points that we have repeatedly emphasized in this class is the importance of visualizing your data, and that process is still important with this type of data. Fortunately there is a great quality profile plot that you can generate with just a single command from `dada2`.

```{r}
# Visualize the quality profile of the first two files containing forward reads

plotQualityProfile(fileNameForwards[[1]])
plotQualityProfile(fileNameForwards[[2]])

```

We see here that the forward reads are really good quality. Callahan advises "trimming the last few nucleotides to avoid less well-controlled errors that can arise there." OTOH, Christopher Taylor, who runs the Metagenomics lab at LSU Health Sciences Center advises to always trim the first 10 reads. 

Let's look at the reverse reads.

```{r}
# Visualize the quality profile of the first two files containing reverse reads

plotQualityProfile(fileNameReverses[[1]])
plotQualityProfile(fileNameReverses[[2]])
```

The quality of the reverse reads is subtantially worse, especially toward the end, a common phenomenon with Illumina paired-end sequencing. The dada algorithm incorporates error quality into the model, so it is robust to lower quality sequences, but trimming is still a good idea.

If you are using your own data, make sure that you have good overlap, the more the better.

#### Performing the Filtering and Trimming

We will use typical filtering parameters.

- `maxN = 0` -- `dada2` requires that there be no N's in a sequence
- `truncQ = 2` -- truncate reads at the first instance of a quality less than or equal to \code{truncQ}#.
- `maxEE` = 2 -- sets the maximum number of expected errors allowed in a read, which is a better filter than simply averaging quality scores.

Let's jointly filter the forward and reverse reads with the fastqPairedFilter function.



```{r}
# Make a directory and filenames for the filtered fastqs

filt.path <- file.path(path, "filtered")
if(!file_test("-d", filt.path)) dir.create(filt.path)
filtForwards <- file.path(filt.path, paste0(sample.names, "_F_filt.fastq.gz"))
filtReverses <- file.path(filt.path, paste0(sample.names, "_R_file.fastq.gz"))

# Now filter

for (i in seq_along(fileNameForwards)) {
  fastqPairedFilter(c(fileNameForwards[i], fileNameReverses[i]), 
                      c(filtForwards[i], filtReverses[i]),
                      truncLen=c(240,160),
                      maxN = 0, maxEE = c(2,2),
                      truncQ = 2, rm.phix=TRUE,
                      compress=TRUE, verbose=TRUE)
}


```

There are other filtering methods, and you can use them. However, for the later `mergePairs` step to work, you need to have the forward and reverse reads in matched order. The fastq files as they come off the Illumina have this property and `fastqPairedFilter` preserves it, but not all other filtering methods do so.

#### Important Note 4

If doing this with data that have not had the primers trimmed off, you can use the `trimleft` argument in the `fastqPairedFilter` call. For instance, if you used the 515F and 805R primers listed above, these are both 19 bp long, so you should trim the first 19 bp from both the forward and the reverse reads. Insert into  the command "trimLeft=c(19,19)", and let the trimming begin!

#### End of Note

### Dereplication

You can gain further efficiencies by dereplicating the reads, ths is combining all identical sequences so that all you are left with is a list of "unique sequences" and a count of them, defined as the "abundance". Other pipelines can do this too to gain efficiency, but `dada2` retains a summary of the quality information associated with each unique sequence, developning a consensus quality profile as the average of the positional qualities from the dereplicated reads, which it then uses to inform the error model in the subsequent denoising step.

```{r}
# Dereplicate

derepForwards <- derepFastq(filtForwards, verbose=TRUE)
derepReverses <- derepFastq(filtReverses, verbose=TRUE)

# Name the derep-class objects by the sample names
names(derepForwards) <- sample.names
names(derepReverses) <- sample.names
```

The above step takes a bit.

#### If using your own data

If you have a a big dataset, get the initial error rate estimates from a subsample of your data.
### Learn the Error Rates

The `dada2` algorithm uses a parametric error model (`err`), and, of course, the amplicon dataset will have different error rates. The algorithm will learn its error model from the data by alternating estimation of error rates and composition of the sample until convergence of the sample on a jointly consistent solution (like the EM algorithm, if you happen to know that) (and if you don't, it does not matter).

So we will run this joint inference 4 times. The first passes will be through the forward and reverse reads setting `selfConsist = TRUE`. The second passes will be through the forward and reverse reads with the learned error structure. On the first pass, the algorithm starts with an initial guess, which is that the maximum possible error rates in these data, that is, the error rates if only the most abundant sequence is correct, and all the rest are errors. This is what happens when we set `err=NULL`.

Let's take a 5 minute break while we take the first pass through the Forward reads then the Reverse reads:

```{r}
dadaForwards.learn <- dada(derepForwards, err=NULL, selfConsist = TRUE, multithread=TRUE)
dadaReverses.learn <- dada(derepReverses, err=NULL, selfConsist = TRUE, multithread=TRUE)
```


Now let's get the initial error estimates:

```{r}

# Store initial error estimates for Forward reads

errForwards <- dadaForwards.learn[[1]]$err_out

# Now for the Reverse reads

errReverses <- dadaReverses.learn[[1]]$err_out

```

Finally it is always worthwhile to visualize the estimated error rates:

```{r}
# Plot the estimated error rates for the Forward reads

plotErrors(dadaForwards.learn[[1]], nominalQ=TRUE)

# And for the Reverse reads

plotErrors(dadaReverses.learn[[1]], nominalQ = TRUE)


```

The error for each possible type of transition (i.e., A -> C, A -> T, ..., T -> G) are shown. The black points are the observed error rates for each consensus quality score. The black line shows the estimated error rates after convergence. The red line is the error rates expected under the nominal definition of the Q value. You see that the black line (estimated rates) fots the observed rates well, and the error rates drop with increased quality as expected. So all is looking good and we proceed.

### Sample Inference

We are now ready to infer the sequence variants in each sample (second dada pass)

```{r}
# First with the Forward reads

dadaForwards <- dada(derepForwards, err = errForwards, multithread = TRUE)

# Then with the Reverse reads

dadaReverses <- dada(derepReverses, err = errReverses, multithread = TRUE)

# Inspect the dada-class objects returned by the dada function

dadaForwards[[1]]
dadaReverses[[1]]

```

We can see that the algorithm has inferred 128 unique sequence variants from the forward reads and 116 from the reverse reads. 

### Merge Paired Reads

We can eliminate further spurious sequence variants by merging overlapping reads. The core function is `mergePairs` and it depends on the forward and reverse reads being in matching order at the time they were dereplicated.

```{r}

# Merge the denoised forward and reverse reads

mergedPairs <- mergePairs(dadaForwards, derepForwards, dadaReverses, derepReverses, verbose = TRUE )

# Inspect the merged data.frame from the first sample

head(mergedPairs[[1]])

```

We now have a `data.frame` object for each sample with the merged `$sequence`, its `abundance`, and the indices of the merged `$forward` and `$reverse` denoised sequences. Pair reads that did not precisely overlap have been removed by the `mergePairs` function.

#### Important Note 5


If doing this with your own data, most of your reads should successfully merge. If this is not the case, you will need to revisit some upstream parameters. In particular, make sure you did not trim away any overlap between reads.

#### End of Note

### Sequence Table Construction

We will now construct the sequence table, this being analogous to the "OTU table" produced by other methods.

```{r}

# Construct sequence table

seqtab <- makeSequenceTable(mergedPairs[names(mergedPairs) != "Mock"])

# Consider the table

dim(seqtab)
class(seqtab)

# Inspect the distribution of sequence lengths

table(nchar(getSequences(seqtab)))


```

We see that the sequence table is a `matrix` with rows corresponding to and named by the samples, and columns corresponding to (and named by) the sequence variants. We also see that the lengths of all of the sequences fall in the range expected for V4 amplicons.

#### Important Note 6

If working with your own data you may find sequences that are much longer or much shorter than expected. These may be the result of non-specific priming, and you should consider removing them. Use the command `seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% seq(250, 256)]`.

#### End of Note

### Remove Chimeras

So far we have let the `dada` function remove substitution errors and indel errors, but chimeras remain. The accuracy of the sequences after denoising makes chimera identification easier than if we had done that earlier with "fuzzier" sequences because all sequences now can be exactly reconstructed as a bimera (two-parent chimera) from more abundant sequences.

```{r}

# Remove chimeric sequences

seqtab.nochim <- removeBimeraDenovo(seqtab, verbose=TRUE)
dim(seqtab.nochim)
table(nchar(getSequences(seqtab.nochim)))
sum(seqtab.nochim)/sum(seqtab)

```

The fraction of chimeras can be substantial. In this example, chimeras account for 63/274 unique sequence variants, or about 23% of them, but these variants account for only about 4% of the total sequence reads.

#### Important Note 7 

Most of the _reads_ should remain after chimera removal, although it is not uncommon for a majority of _sequence variants_ to be removed. If most of your reads are removed as chimeric, you may need to revisit upstream processing. In almost all cases this is caused by primer sequences with ambiguous nucleotides that were not removed prior to beginning the DADA2 pipeline.

#### End of Note

### Assign Taxonomy

Most people want to know the names of the organisms associated with the sequence variants, and so we want to classify them taxonomically. The package will use a classifier for this purpose, taking a set of sequences and a training set of taxonomically classified sequences, and outputs taxonomic assignments with at least `minBoot` bootstrap confidence. 

There are many training sets to use. GreenGenes is one such set, but it has not been updated in 3 years. SILVA and UNITED ITS are others, the latter being used for fungi. We are going to use a training set from the Ribosomal Database Project. You should have downloaded that earlier and it should be sitting in the same folder as the original forward and reverse read files.

```{r}

# Assign taxonomy

# First initialize random number generator for reproducibility

set.seed(100)
getwd()
path
list.files(path)
reference_fasta <- "~/Documents/NRSG_741/MiSeqData/MiSeq_SOP/rdp_train_set_14.fa.gz"
taxa <- assignTaxonomy(seqtab.nochim, refFasta = reference_fasta)
unname(head(taxa))
```

### Species Assignment

We can also use the RDP species assignment dataset to do exactly that, that is, to assign species.

```{r}

# Assign species

genus.species <- assignSpecies(seqtab.nochim, "~/Documents/NRSG_741/MiSeqData/MiSeq_SOP/rdp_species_assignment_14.fa.gz")

```

### Construct a Phylogenetic Tree

Phylogenetic relatedness is often used to inform downstream analyses, particularly the calculation of phylogeny-aware distances between microbial communities, including the weighted and unweighted UniFrac distances. We can use the reference-free `dada2` sequence inference to construct the phylogenetic tree by relating the inferred sequence variants *de novo*. First we will need to install the `DECIPHER` package from Bioconductor and the `phangorn` package from CRAN.

Next call them up then we will use DECIPHER for multiple sequence alignment

```{r}
library(DECIPHER)
seqs <- getSequences(seqtab.nochim)

# This next command will allow propagation of sequence names to the tip labels of the tree
names(seqs) <- seqs
alignment <- AlignSeqs(DNAStringSet(seqs), anchor=NA)
```

Now that the sequences are aligned we can use the `phanghorn` package to construct the tree. 

```{r}

library(phangorn)

# Construct the tree
phang.align <- phyDat(as(alignment, "matrix"), type="DNA")
dm <- dist.ml(phang.align)
treeNJ <- NJ(dm) # Tip order will not equal sequence order
fit <- pml(treeNJ, data=phang.align)

## negative edges length changed to 0.

fitGTR <- update(fit, k=4, inv=0.2)
fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE, 
                    rearrangement = "stochastic", control=pml.control(trace=0))
detach("package:phangorn", unload=TRUE)

```

We will be viewing the tree below.

### Handoff to `phyloseq`

Our next activity will be to hand off the data to the `phyloseq` package for analysis. This package requires three items: the "OTUtable," the taxonomy table, and data about the samples. The first two items are directly available at the end of your `dada2`run, and you can import the latter as a .csv file. In the case of the data that are considered here, we can calculate the derive the gender (G), mouse subject number (X), and day post-weaning (Y) directly from the file name, which has the form GXDY.

```{r}
# Create a data frame for the sample data
samples.out <- rownames(seqtab.nochim)

# Create subject, gender, and day variables
subject <- sapply(strsplit(samples.out, "D"), `[`, 1)
gender <- substr(subject,1,1)
subject <- substr(subject, 2, 999)
day <- as.integer(sapply(strsplit(samples.out, "D"), `[`, 2))

# Combine into dataframe
samdf <- data.frame(Subject = subject, Gender = gender, Day = day)

#Create indicator of early or late day of post-weaning
samdf$When <- "Early"
samdf$When[samdf$Day > 100] <- "Late"

# Assign rownames to the dataframe == these will be the same as the rownames of the "OTUtable"
rownames(samdf) <- samples.out
```

Now that we have our sample data, let's create the phyloseq object.

```{r}
library(phyloseq)

# Create phyloseq object
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf),
               tax_table(taxa))

# Describe it
ps
```

